---
title: "Car Price Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Car Price Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/deepcontractor/car-price-prediction-challenge\>*

### Reference:

*\<deepcontractor. (n.d.). Car Price Prediction Challenge [Dataset]. Kaggle. https://www.kaggle.com/datasets/deepcontractor/car-price-prediction-challenge\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Load dataset
```{r Load dataset}
# Load dataset
car_data <- read.csv("car_price_prediction.csv", colClasses = c(
  Levy = "numeric",
  Manufacturer = "character",
  Model = "character",
  Category = "factor",
  Leather_interior = "factor",
  Fuel_type = "factor",
  Engine_volume = "numeric",
  Mileage = "numeric",
  Cylinders = "numeric",
  Gear_box_type = "factor",
  Drive_wheels = "factor",
  Wheel = "factor",
  Color = "factor",
  Airbags = "numeric",
  Price = "numeric"
))

# Display the structure of the dataset
str(car_data)

# View the first few rows of the dataset
head(car_data)

# View the dataset in a separate viewer window
View(car_data)
```

## Measures of Frequency
```{r MOF}
# Load the required libraries
library(dplyr)
library(e1071)

# Measures of Frequency
frequency_count <- car_data %>%
  select(Manufacturer, Model, Category, Leather_interior, Fuel_type, Gear_box_type, Drive_wheels, Wheel, Color) %>%
  summarise_all(~list(table(.)))

# Display frequency counts
frequency_count
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
central_tendency <- car_data %>%
  summarise(
    mean_Levy = mean(Levy, na.rm = TRUE),
    median_Levy = median(Levy, na.rm = TRUE),
    mean_Engine_volume = mean(Engine_volume, na.rm = TRUE),
    median_Engine_volume = median(Engine_volume, na.rm = TRUE),
    mean_Mileage = mean(Mileage, na.rm = TRUE),
    median_Mileage = median(Mileage, na.rm = TRUE),
    mean_Cylinders = mean(Cylinders, na.rm = TRUE),
    median_Cylinders = median(Cylinders, na.rm = TRUE),
    mean_Airbags = mean(Airbags, na.rm = TRUE),
    median_Airbags = median(Airbags, na.rm = TRUE),
    mean_Price = mean(Price, na.rm = TRUE),
    median_Price = median(Price, na.rm = TRUE)
  )

# Display measures of central tendency
central_tendency
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
distribution_measures <- car_data %>%
  summarise(
    var_Levy = var(Levy, na.rm = TRUE),
    sd_Levy = sd(Levy, na.rm = TRUE),
    range_Levy = diff(range(Levy, na.rm = TRUE)),
    skewness_Levy = skewness(Levy, na.rm = TRUE),
    kurtosis_Levy = kurtosis(Levy, na.rm = TRUE),
    
    var_Engine_volume = var(Engine_volume, na.rm = TRUE),
    sd_Engine_volume = sd(Engine_volume, na.rm = TRUE),
    range_Engine_volume = diff(range(Engine_volume, na.rm = TRUE)),
    skewness_Engine_volume = skewness(Engine_volume, na.rm = TRUE),
    kurtosis_Engine_volume = kurtosis(Engine_volume, na.rm = TRUE),
    
    var_Mileage = var(Mileage, na.rm = TRUE),
    sd_Mileage = sd(Mileage, na.rm = TRUE),
    range_Mileage = diff(range(Mileage, na.rm = TRUE)),
    skewness_Mileage = skewness(Mileage, na.rm = TRUE),
    kurtosis_Mileage = kurtosis(Mileage, na.rm = TRUE),
    
    var_Cylinders = var(Cylinders, na.rm = TRUE),
    sd_Cylinders = sd(Cylinders, na.rm = TRUE),
    range_Cylinders = diff(range(Cylinders, na.rm = TRUE)),
    skewness_Cylinders = skewness(Cylinders, na.rm = TRUE),
    kurtosis_Cylinders = kurtosis(Cylinders, na.rm = TRUE),
    
    var_Airbags = var(Airbags, na.rm = TRUE),
    sd_Airbags = sd(Airbags, na.rm = TRUE),
    range_Airbags = diff(range(Airbags, na.rm = TRUE)),
    skewness_Airbags = skewness(Airbags, na.rm = TRUE),
    kurtosis_Airbags = kurtosis(Airbags, na.rm = TRUE),
    
    var_Price = var(Price, na.rm = TRUE),
    sd_Price = sd(Price, na.rm = TRUE),
    range_Price = diff(range(Price, na.rm = TRUE)),
    skewness_Price = skewness(Price, na.rm = TRUE),
    kurtosis_Price = kurtosis(Price, na.rm = TRUE)
  )

# Display measures of distribution
distribution_measures   
```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numeric columns
correlation_matrix <- cor(car_data %>% select(Levy, Engine_volume, Mileage, Cylinders, Airbags, Price), use = "complete.obs")

# Covariance matrix for numeric columns
covariance_matrix <- cov(car_data %>% select(Levy, Engine_volume, Mileage, Cylinders, Airbags, Price), use = "complete.obs")

# Display correlation and covariance matrices
correlation_matrix
covariance_matrix
```

## ANOVA
```{r ANOVA}
# Load the required library
library(car)

# One-way ANOVA for Category
anova_category <- aov(Price ~ Category, data = car_data)
summary(anova_category)

# One-way ANOVA for Fuel_type
anova_fuel_type <- aov(Price ~ Fuel_type, data = car_data)
summary(anova_fuel_type)

# One-way ANOVA for Gear_box_type
anova_gear_box_type <- aov(Price ~ Gear_box_type, data = car_data)
summary(anova_gear_box_type)

# Post-hoc test (Tukey's HSD) to see which specific groups differ if ANOVA is significant
tukey_category <- TukeyHSD(anova_category)
tukey_fuel_type <- TukeyHSD(anova_fuel_type)
tukey_gear_box_type <- TukeyHSD(anova_gear_box_type)

# Display post-hoc test results
tukey_category
tukey_fuel_type
tukey_gear_box_type
```

## Plots
```{r Plots}
# Load the required libraries
library(ggplot2)
library(GGally)

# Univariate Plots

# Histogram for numeric variables
ggplot(car_data, aes(x = Price)) + 
  geom_histogram(binwidth = 1000, fill = "blue", color = "black") +
  labs(title = "Histogram of Car Prices", x = "Price", y = "Frequency")

ggplot(car_data, aes(x = Mileage)) + 
  geom_histogram(binwidth = 10000, fill = "green", color = "black") +
  labs(title = "Histogram of Car Mileages", x = "Mileage", y = "Frequency")

# Box plot for numeric variables
ggplot(car_data, aes(x = Category, y = Price)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Box Plot of Car Prices by Category", x = "Category", y = "Price")

# Bar chart for categorical variables
ggplot(car_data, aes(x = Fuel_type)) + 
  geom_bar(fill = "purple") +
  labs(title = "Bar Chart of Fuel Types", x = "Fuel Type", y = "Count")

# Multivariate Plots

# Scatter plot for numeric variables
ggplot(car_data, aes(x = Mileage, y = Price)) + 
  geom_point(color = "red") +
  labs(title = "Scatter Plot of Price vs Mileage", x = "Mileage", y = "Price")

# Scatter plot with a categorical variable
ggplot(car_data, aes(x = Mileage, y = Price, color = Fuel_type)) + 
  geom_point() +
  labs(title = "Scatter Plot of Price vs Mileage by Fuel Type", x = "Mileage", y = "Price")

# Pair plot for numeric variables
ggpairs(car_data %>% select(Levy, Engine_volume, Mileage, Cylinders, Airbags, Price), 
        title = "Pair Plot of Numeric Variables")

# Box plot by a categorical variable
ggplot(car_data, aes(x = Gear_box_type, y = Price, fill = Gear_box_type)) +
  geom_boxplot() +
  labs(title = "Box Plot of Car Prices by Gear Box Type", x = "Gear Box Type", y = "Price")

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Summary of missing values
missing_values_summary <- sapply(car_data, function(x) sum(is.na(x)))
missing_values_summary

# Visualize missing values using a heatmap
library(naniar)
library(ggplot2)

# Plot missing values
gg_miss_var(car_data) + labs(title = "Missing Values in Car Dataset")

# Detailed plot of missing data
vis_miss(car_data) + labs(title = "Detailed Missing Values in Car Dataset")

```

## Omit Rows
```{r Omit Rows}
# Remove rows with missing values
car_data_clean <- na.omit(car_data)

# Summary of missing values after removing rows
missing_values_after <- sapply(car_data_clean, function(x) sum(is.na(x)))
print("Missing Values After Removing Rows:")
print(missing_values_after)
```

# Training Model
## Data Splitting
```{r Data Splitting}
library(caTools)

# Split the cleaned dataset into training and testing sets
set.seed(123)  # for reproducibility
split <- sample.split(car_data_clean$Price, SplitRatio = 0.7)
train_data <- subset(car_data_clean, split == TRUE)
test_data <- subset(car_data_clean, split == FALSE)

# Display the dimensions of the training and testing sets
print("Training Set Dimensions:")
print(dim(train_data))
print("Testing Set Dimensions:")
print(dim(test_data))
```

## Bootstrapping
```{r Bootstrapping}
# Perform bootstrapping on the cleaned dataset
boot_samples <- 1000
bootstrapped_means <- numeric(boot_samples)

for (i in 1:boot_samples) {
  boot_sample <- sample(car_data_clean$Price, replace = TRUE)
  bootstrapped_means[i] <- mean(boot_sample)
}

# Summary of bootstrapped means
summary_boot_means <- summary(bootstrapped_means)
print("Summary of Bootstrapped Means:")
print(summary_boot_means)

# Plot histogram of bootstrapped means
hist(bootstrapped_means, main = "Bootstrapped Means", xlab = "Mean Price")
```

## Cross-validation
```{r Cross-validation}
# Load the required library
library(caret)

# Define training control
ctrl <- trainControl(method = "cv", number = 5)

# Perform cross-validation with linear regression on the cleaned dataset
lm_model <- train(Price ~ ., data = train_data, method = "lm", trControl = ctrl)

# Print the cross-validated linear regression model
print(lm_model)
```

## Training Different Models
```{r Model Training}
# Load the required library
library(glmnet)

# Prepare data
x_train <- as.matrix(train_data[, -15])  # Exclude the target variable (Price)
y_train <- train_data$Price

# Train Lasso regression model
lasso_model <- cv.glmnet(x_train, y_train, alpha = 1)

# Display the best lambda value
print("Best Lambda for Lasso:")
print(lasso_model$lambda.min)

# Display the coefficients of the Lasso model
print("Coefficients of the Lasso model:")
print(coef(lasso_model, s = "lambda.min"))

# Train Ridge regression model
ridge_model <- cv.glmnet(x_train, y_train, alpha = 0)

# Display the best lambda value
print("Best Lambda for Ridge:")
print(ridge_model$lambda.min)

# Display the coefficients of the Ridge model
print("Coefficients of the Ridge model:")
print(coef(ridge_model, s = "lambda.min"))

print(ridge_model)
```

## Performance Comparison
```{r Performance comparison}
# Load the required libraries
library(caret)
library(glmnet)

# Prepare the data
x <- as.matrix(car_data_clean[, -15])  # Features (excluding the target variable)
y <- car_data_clean$Price  # Target variable

# Define the training control
ctrl <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

# Define the models
models <- c("lm_model", "lasso_model", "ridge_model")

# Define model parameters
model_params <- list(
  lm_model = list(method = "lm"),
  lasso_model = list(method = "glmnet", alpha = 1, tuneGrid = expand.grid(lambda = seq(0.001, 1, length = 100))),
  ridge_model = list(method = "glmnet", alpha = 0, tuneGrid = expand.grid(lambda = seq(0.001, 1, length = 100)))
)

# Train the models
res <- lapply(models, function(model_name) {
  train(x, y, method = model_params[[model_name]]$method, trControl = ctrl, tuneGrid = model_params[[model_name]]$tuneGrid)
})

# Compare model performances
resamples <- resamples(res)
summary(resamples)
```

# Saving Model
```{r Saving Model}
# Save the trained Lasso model to a file
saveRDS(lasso_model, file = "./models/lasso_model.rds")

# Load the saved Lasso model
loaded_model_lasso <- readRDS("./models/lasso_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Levy = 1399,
  Manufacturer = "LEXUS",
  Model = "RX 450",
  Category = "Jeep",
  Leather_interior = "Yes",
  Fuel_type = "Hybrid",
  Engine_volume = 3.5,
  Mileage = 186005,
  Cylinders = 6,
  Gear_box_type = "Automatic",
  Drive_wheels = "4x4",
  Wheel = "Left wheel",
  Color = "Silver",
  Airbags = 12
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_model_lasso, s = "lambda.min", newx = as.matrix(new_data))

# Print predictions
print(predictions_loaded_model)

```


